---
title: "Relatedness of GSMNP pigs"
author: "Matt DeSaix"
date: "2024-09-11"
output: html_document
---


Going to use Courtney's method of identifying relatedness in the presence of population structure (with PCrelate/PCair) on the Missouri pigs. It will be interesting to see too the difference in estimates compared to KING-robust


```{r, message = FALSE}
library(SNPRelate)
library(GENESIS)
library(GWASTools)
```


```{r}
#Specify genotype file prefix
Genotypes <- "./data/genotypes/buffer-10k/grsm-10k-final-571_10508"

#Create .gds file using .bed .bim .fam
snpgdsBED2GDS(bed.fn = paste0(Genotypes,".bed"),
              bim.fn = paste0(Genotypes,".bim"),
              fam.fn = paste0(Genotypes,".fam"),
              out.gdsfn = paste0(Genotypes,".gds"))
```



Calculate IBD coefficients by KING method of moment using gds
"KING-robust" -- relationship inference in the presence of population stratification

```{r}
#Open GDS
gds <- snpgdsOpen(paste0(Genotypes,".gds"))
##KING-robust is the default if no type is specified 
king <- snpgdsIBDKING(gds, type="KING-robust")
#create a table of kinship coefficients. Could write this table out if desired. 
KINGtab <- snpgdsIBDSelection(king)
#create matrix of KING relationship estimates
KINGmat <- kingToMatrix(king)
snpgdsClose(gds)
```

```{r}
outking <- paste0("./out/pcair-relate/buffer-10k/GSMNP-KING-Kinship.csv")
data.table::fwrite(KINGtab, file = outking, 
       col.names = T, row.names=F,quote=F, sep=",")
```

```{r}
dim(KINGmat)
```


```{r}
## --------------------------------------------------------------------------
#read in GDS data created above
HapMap_geno <- GdsGenotypeReader(paste0(Genotypes,".gds"))
# create a genotype data class object
HapMap_genoData <- GenotypeData(HapMap_geno)

## --------------------------------------------------------------------------
# run PC-AiR on pruned SNPs using KING relationship matrix and genotype data read from .gds. 
##Generates principal components
###Change eigen.cnt to control the number of eigenvectors returned. If <= 0 to then return all eigenvectors (Number of eigenvectors = number of unrelated individuals)

mypcair <- pcair(HapMap_genoData, kinobj = KINGmat, divobj = KINGmat, eigen.cnt = 0)
```

```{r}
summary(mypcair)
```


```{r}
#Save variance proportions, eigenvalues, etc from PC-AiR PCA.

Varprop <- as.data.frame(mypcair$varprop)
Eigenvalues <- as.data.frame(mypcair$values)
Eigenvectors <- as.data.frame(mypcair$vectors)

outprefix <- paste0("./out/pcair-relate/buffer-10k/GSMNP-PCAiR-", ncol(Eigenvectors), "PCs-")
# write.csv(Varprop, paste0(outprefix, "VarProportions.csv"), quote = FALSE, row.names = TRUE)
# write.csv(Eigenvalues, paste0(outprefix, "Eigenvalues.csv"), quote = FALSE, row.names = TRUE)
# write.csv(Eigenvectors, paste0(outprefix, "Eigenvectors.csv"), quote = FALSE, row.names = TRUE)

## --------------------------------------------------------------------------
#Generate scree plot of eigenvalues to evaluate # of PC to include in PC-Relate

#read function
# source("./scripts/ScreePlot_Function.R")
# 
# scree.plot(mypcair$values[1:ncol(Eigenvectors)], lab = FALSE, 
#            title = "GRSM PC-AiR", 
#            new.plot = TRUE, pdf.file = TRUE, 
#            pdf.name = "./out/pcair-relate/buffer-10k/GSMNP-PCAiR-Screeplot.pdf")
```


```{r}
#PCA plots
plot(mypcair)
plot(mypcair, vx = 1, vy = 2)
```


## Use automatic PC selection tool - from Courtney's relatedness work

```{r}
library(PCAtools)
library(mousetRajectory)
library(findPC)
```

```{r}
pcaireigenvalues <- setNames(mypcair$values,seq(1,ncol(Eigenvectors),1))

is_monotonic(pcaireigenvalues, decreasing = T)

NumPCsRetain <- findElbowPoint(pcaireigenvalues)
NumPCsRetain
```

`findPC` method

```{r}
pcairsdev <- sqrt(mypcair$values)
# NumPCsRetain2 <- findPC(sdev=pcairsdev,number=length(pcairsdev),method="perpendicular line",figure=T)[,1]
NumPCsRetain2 <- findPC(sdev=pcairsdev,number=length(pcairsdev),method="perpendicular line",figure=T)
```

                     
```{r}
# run PC-Relate to obtain relationship matrix 
HapMap_genoData <- GenotypeBlockIterator(HapMap_genoData)

#pcs = A matrix of principal components (PCs) to be used for ancestry adjustment.
##Each column represents a PC, and each row represents an individual
##When drop is FALSE, the dimensions of the object are kept.

#training.set = vector of individual IDs specifying which samples are used to estimate the ancestry adjustment at each SNP

# small.samp.correct = Post-hoc correction to try to prevent over adjustment driven by a small number of samples with high leverage due to unique ancestry. Recommended if < 5,000 samples in your analysis (Conomos even used it up to ~10,000 samples)

# run pcrelate() with PCs 1 through i
mypcrel <- pcrelate(HapMap_genoData, 
                     pcs = mypcair$vectors[, 1:NumPCsRetain, drop = FALSE],
                     training.set = mypcair$unrels,
                     small.samp.correct = TRUE,
                     verbose = TRUE)
```
  
  
  
  
```{r}
# Convert PC-Relate results to a matrix 
grm <- pcrelateToMatrix(mypcrel, scaleKin = 1) # specifies a factor to multiply the kinship coefficients by in the GRM 
  
#Convert to data frame so that GRM can be saved
matrix <- as.data.frame(as.matrix(grm))
# outname <- paste0("./out/pcair-relate/buffer-10k/GSMNP-PCRelate-RelatednessMatrix-", NumPCsRetain, "PCs.csv")
# write.csv(matrix, outname)
  
#OR
```

```{r}
#Output results in long format to extract kinship and k(0) for simulated individuals
kinBtwn <- mypcrel[["kinBtwn"]]
  
#Include header because files will not be merged
outname2 <- paste0("./out/pcair-relate/buffer-10k/GSMNP-PCRelate-Kinship-", NumPCsRetain, "PCs.csv")
# data.table::fwrite(kinBtwn, file = outname2, 
#        col.names = T, row.names=F,quote=F, sep=",")
 
```


## Compare KING vs PCrelate


```{r, message = FALSE, warning = FALSE}
library(tidyverse)
```

```{r}
full.kin <- KINGtab %>%
  select(ID1, ID2, kinship) %>%
  rename("KING" = "kinship") %>%
  left_join(kinBtwn %>%
              select(ID1, ID2, kin) %>%
              rename("PCRelate" = "kin"),
            by = c("ID1" = "ID1", 
                   "ID2" = "ID2"))
head(full.kin)
```

```{r}
summary(full.kin)
```

Use the kinship cutoff of 0.177 (i.e. 1st and 2nd degree relatives)

```{r}
full.kin %>%
  filter(PCRelate >= 0.177) %>%
  group_by(ID1) %>%
  summarize(N = n()) %>%
  arrange(desc(N))
```



Organize the data to get an ordered list of individuals based on how many other related individuals they have...thus, can prioritize removing individuals who have the most relatives

```{r}
full.kin.related.ID1 <- full.kin %>%
  filter(PCRelate >= 0.177) %>%
  group_by(ID1) %>%
  summarize(N = n(),
            .groups = "drop") %>%
  rename("ID" = "ID1")

full.kin.related.ID2 <- full.kin %>%
  filter(PCRelate >= 0.177) %>%
  group_by(ID2) %>%
  summarize(N = n(),
            .groups = "drop") %>%
  rename("ID" = "ID2")

full.kin.related.tot <- rbind(full.kin.related.ID1,
                              full.kin.related.ID2) %>%
  group_by(ID) %>%
  summarize(Total = sum(N),
            .groups = "drop") %>%
  arrange(desc(Total)) %>%
  pull(ID)
```

Below, create a vector of the individual IDs to remove based on relatedness. Iteratively, go through the names of related individuals (from above - ordered by individuals with most relatives), and remove an individual and add that to the list until there are no more related pairs.

```{r}
related.remove <- c()

tmp.related.tib <- full.kin %>%
  filter(PCRelate >= 0.177)

for(i in 1:length(full.kin.related.tot)){
  tmp.ind <- full.kin.related.tot[i]
  tmp.related.tib.new <- tmp.related.tib %>%
    filter(ID1 != tmp.ind,
           ID2 != tmp.ind)
  if(nrow(tmp.related.tib) != nrow(tmp.related.tib.new)){
    related.remove <- c(related.remove, tmp.ind)
  }
  if(nrow(tmp.related.tib.new) == 0){
    print("All done!")
    break
  }
  tmp.related.tib <- tmp.related.tib.new
  
}
writeLines(related.remove,
           "./data/relatedness/buffer-10k/grsm-10k-related-pig-IDs.txt")

length(related.remove)
```



```{r}
full.kin %>%
  filter(KING > 0) %>%
  ggplot() +
  geom_point(aes(x = KING,
                 y = PCRelate))
```


```{r}
full.kin %>%
  ggplot() +
  geom_point(aes(x = KING,
                 y = PCRelate))
```

### Create unrelated "keep" file for PLINK

```{r}
filtered.fam <- read_table("./data/genotypes/buffer-10k/grsm-10k-final-571_10508.fam",
                           col_names = FALSE)
unrelated.keep <- filtered.fam %>%
  select(1,2) %>%
  filter(!X2 %in% related.remove)
write_delim(unrelated.keep, "./data/relatedness/buffer-10k/grsm-10k-unrelated-keep.txt",
            delim = "\t", col_names = FALSE)
unrelated.keep
```

```{sh}
plink=../software/plink-1.9/plink
bfile=./data/genotypes/buffer-10k/grsm-10k-final-571_10508
keep=./data/relatedness/buffer-10k/grsm-10k-unrelated-keep.txt
outname=./data/genotypes/buffer-10k/grsm-10k-final-unrelated-282_10508
${plink} --noweb --bfile ${bfile} --keep ${keep} --make-bed --recode A --out ${outname}
```




