---
title: "Population structure in the Smokies"
author: "Matt DeSaix"
date: "2024-09-11"
output: html_document
---

```{r, message = FALSE}
# BiocManager::install("LEA")
library(LEA)
library(randomcoloR)
library(tidyverse)
```

Running through population structure analyses based on Rachael's workflow (and code!)


First off, convert data from raw PLINK to appropriate for LEA package

```{r}
LEA_geno <- read.table("./data/genotypes/buffer-10k/grsm-10k-final-571_10508.raw", header=T)
LEA_geno <- LEA_geno[,7:ncol(LEA_geno)]
LEA_geno <- sapply( LEA_geno, as.numeric )
LEA_geno[is.na(LEA_geno)] <- 9 #replace missing data with 9s
geno.name <- "./out/snmf/grsm-10k-final-571_10508.012.geno"
write.geno(LEA_geno, geno.name)
```

Now to run through snmf with different K values

```{r}
#Run LEA
max_k <- 8

project <- NULL
project <- snmf(geno.name,
               K = 1:max_k,
               entropy = TRUE,
               repetitions = 10,
               alpha = 1000,
               project = "new",
               CPU = 4)
```

Identify repititions with lowest cross-entropy for each K

```{r}
# Plink_Keep <- read.delim("./data/relatedness/GSMNP-unrelated-keep.txt", header=F)
# colnames(Plink_Keep) <- c("Fam","SubjectID")
Plink_Keep <- read.delim("./data/genotypes/buffer-10k/grsm-10k-final-571_10508.fam", header = F, sep = " ")[,c(1,2)]
colnames(Plink_Keep) <- c("Fam","SubjectID")

#Load existing LEA project
project <- load.snmfProject("./out/snmf/grsm-10k-final-571_10508.012.snmfProject")

num_K <- unique(project@K)

best_K <- 1:length(num_K)
  
for(i in 1:length(best_K)){
  best_K[i]<-which.min(cross.entropy(project, K=i))
}
```

Plot cross entropy values

```{r}
plot(project, col = "blue", pch = 19, cex = 1.2)
```


```{r}
pdf("./plots/population_structure/snmf/GRSM-10k_PopStructurePlots_alpha1000.pdf")
plot(project, col = "blue", pch = 19, cex = 1.2) # Lower values means better model
  
col_pop<-distinctColorPalette(k=max(num_K))
  
#best_K
  
for(i in 2:length(num_K)){
    barchart(project, K = num_K[i], run = best_K[i],
             border = NA, space = 0,
             xlab = "Individuals",
             ylab = "Ancestry proportions",
             col=col_pop,
             main = paste0("K= ",i),
             sort.by.Q=TRUE) -> bp
    axis(1, at = 1:length(bp$order),
         labels = Plink_Keep$SubjectID[bp$order], las=3,
         cex.axis = .4)
}
dev.off()

```

```{r}
meta <- read_csv("./data/genotypes/buffer-10k/grsm-10k-meta.csv",
                 show_col_types = FALSE)
sample.meta <- Plink_Keep %>%
  select(SubjectID) %>%
  left_join(meta, by = "SubjectID")
```

```{r}
# gsmnp_sample_coords <- sample.meta %>%
#   select(Long, Lat)
# write.table(gsmnp_sample_coords, file = "./data/feems/grsm.10k.sample.coord",
#             col.names = FALSE, row.names = FALSE)
############ Assign pop IDs to all individuals ##############
# LEAids <- Plink_Keep$SubjectID
#   
# for(i in 2:length(best_K)){
#   LEA_Q <- Q(project, K=i, run=best_K[i])
#   colnames(LEA_Q) <- 1:ncol(LEA_Q)
#   LEA_K <- colnames(LEA_Q)[max.col(LEA_Q,ties.method="first")]
#     
#   temp_PopData <- cbind(LEAids,LEA_K)
#   colnames(temp_PopData)[1] <- "SubjectID"
#   sample.meta <- merge(sample.meta,temp_PopData, by="SubjectID")
#   colnames(sample.meta)[ncol(sample.meta)] <- paste0("K",i)
# }
# write.csv(sample.meta, file="./data/GSMNP_Meta_1Oct2024_popStruct_alpha100.csv")
```



# Plot some K values

## Plot K=2

```{r}
ce_k2 <- cross.entropy(project, K=2)
best_run_k2 <- which.min(ce_k2)
q_mat_k2 <- LEA::Q(project, K = 2, run = best_run_k2)
colnames(q_mat_k2) <- paste0("K", 1:2)
q_df_arrange_k2 <- sample.meta %>%
  cbind(q_mat_k2) %>%
  arrange(desc(K1))
write_csv(q_df_arrange_k2, "./out/snmf/q_df_arrange_k2.csv")


q_df_arrange_k2
```

```{r}
q_df_arrange_k2 %>%
  pivot_longer(cols = K1:K2,
               names_to = "K",
               values_to = "Q") %>%
  mutate(SubjectID = factor(SubjectID, levels = q_df_arrange_k2$SubjectID)) %>% 
  ggplot() +
  geom_col(aes(x = SubjectID, y = Q, fill = K)) +
  theme_minimal()
```

```{r}
q_df_arrange_k2 %>%
  pivot_longer(cols = K1:K2,
               names_to = "K",
               values_to = "Q") %>%
  group_by(SubjectID) %>%
  filter(Q == max(Q)) %>%
  ggplot() +
  geom_point(aes(x = Long, y = Lat, color = K, alpha = Q),
             size = 4) +
  theme_minimal() +
  coord_equal()
```


*Basic piechart with PieGlyph*
plot <- map + 
  # Add pie-charts for each state
  geom_pie_glyph(aes(y = pie_lat, x = pie_long),
                 data = votes_data, colour = 'black',
                 slices = c('Democratic','Republic','Other'))+
  # Colours of the pie sectors
  scale_fill_manual(values = c('#047db7','#c52d25', 'grey'), name = 'Party')+
  # Place legend on top of the plot
  theme(legend.position = 'top')
plot

```{r}
library(PieGlyph)
library(RColorBrewer)
library(sf)
cluster_cols <- brewer.pal(n = 5, name = "Dark2")

gsmnp_boundary <- st_read("./data/shapefiles/GSMNP-park-boundary2/GRSM_BOUNDARY_POLYGON.shp") %>%
  st_transform(4326)
gsmnp_boundary$ID <- factor(seq.int(nrow(gsmnp_boundary)))
gsmnp_core <- gsmnp_boundary %>%
  filter(ID == 17)

p.k2 <- q_df_arrange_k2 %>%
  ggplot() +
  geom_sf(data = gsmnp_core) +
  geom_pie_glyph(aes(y = Lat, x = Long),
                 radius = 0.25,
                 slice = c("K1", "K2"),
                 colour = "grey20") +
  theme_classic() +
  scale_fill_manual(name = "Genetic cluster",
                    values = cluster_cols[c(1,2)],
                    breaks = c("K1", "K2")) +
  xlab("Longitude") +
  ylab("Latitude") +
  guides(fill = guide_legend(override.aes = list(fill = cluster_cols[c(2,1)])))

p.k2
```

```{r}
ind_long_arrange <- q_df_arrange_k2 %>%
  arrange(Long)
p.bar.k2 <- q_df_arrange_k2 %>%
  pivot_longer(cols = K1:K2,
               names_to = "K",
               values_to = "Q") %>%
  mutate(SubjectID = factor(SubjectID, levels = ind_long_arrange$SubjectID)) %>% 
  ggplot() +
  geom_col(aes(x = SubjectID, y = Q, fill = K)) +
  scale_fill_manual(name = "Genetic cluster",
                    values = cluster_cols[c(1,2)]) +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.x = element_blank()) +
  guides(fill = guide_legend(override.aes = list(fill = cluster_cols[c(2,1)])))

ggsave(filename = "./plots/population_structure/snmf/p.k2.bar.ind_long.png",
       plot = p.bar.k2)
p.bar.k2
```


## Plot K=3

```{r}
source("./scripts/gsmnp-functions.R")
k3_plots <- get_snmf_plots(project = project,
                           choiceK = 3,
                           SMsub = sample.meta)

ce_k3 <- cross.entropy(project, K=3)
best_run_k3 <- which.min(ce_k3)
q_mat_k3 <- LEA::Q(project, K = 3, run = best_run_k3)
colnames(q_mat_k3) <- paste0("K", 1:3)
q_df_arrange_k3 <- sample.meta %>%
  cbind(q_mat_k3)
write_csv(q_df_arrange_k3, "./out/snmf/q_df_arrange_k3.csv")
```


## Plot K=4

```{r}
k4_plots <- get_snmf_plots(project = project,
                           choiceK = 4,
                           SMsub = sample.meta)
ce_k4 <- cross.entropy(project, K=4)
best_run_k4 <- which.min(ce_k4)
q_mat_k4 <- LEA::Q(project, K = 4, run = best_run_k4)
colnames(q_mat_k4) <- paste0("K", 1:4)
q_df_arrange_k4 <- sample.meta %>%
  cbind(q_mat_k4)
write_csv(q_df_arrange_k4, "./out/snmf/q_df_arrange_k4.csv")
